---
title: "laurabw_OriginalHomeworkCode_02"
author: "Laura Brubaker-Wittman"
date: "9/24/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem:

Every Saturday, at the same time, a primatologist goes and sits in the forest in the morning and listens for titi
monkey calls, counting the number of calls they hear in a 2-hour window from 5am to 7am. Based on previous
knowledge, she believes that the mean number calls she will hear in that time is exactly 15. Let ***X*** represent the appropriate Poisson random variable of the number of calls heard in each monitoring session.

**Questions:**

1. *What is the probability that she will hear more than 8 calls during any given session?*

    For this one, I will use the ppois() function, since we want to know the cumulative probability of 9 - 15
    calls.

```{r}
x <- 9:15 # The specific interval we are looking for probabilities for
l <- 15 # How many we expect in an hour
ppois(q = x, l)
```

    So, based on this, the probability of hearing 9 or more titi calls is 56.8%

2. *What is the probability that she will hear no calls in a session?*

    Not sure how this one will work? But let's try using the same formula:

```{r}
x <- 0
l <- 15
ppois(q = 0, l)
```

*Yikes!* So basically a **very** small chance of hearing 0 . That makes sense!
    
3. *What is the probability that she will hear exactly 3 calls in a session?*

    Well, it seemed to work with setting it for just ) in the last question, let's see what happens when we set x
    at 3 and not a range like in question 1.
    
```{r}
x <- 3
l <- 15
ppois(q = x, l)
```

Ok, still a *really* number. While if the average is 15 and 3 is way below that, so a small probability for
exactly 3 makes sense, I am not sure if this is how we sent up the function for this type of question.