---
title: "laurabw_OriginalHomeworkCode_02"
author: "Laura Brubaker-Wittman"
date: "9/24/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem:

Every Saturday, at the same time, a primatologist goes and sits in the forest in the morning and listens for titi
monkey calls, counting the number of calls they hear in a 2-hour window from 5am to 7am. Based on previous
knowledge, she believes that the mean number calls she will hear in that time is exactly 15. Let ***X*** represent the appropriate Poisson random variable of the number of calls heard in each monitoring session.

**Questions:**

1. *What is the probability that she will hear more than 8 calls during any given session?*

    For this one, I will use the ppois() function, since we want to know the cumulative probability of 9 - 15
    calls.

```{r}
x <- 9:15 # The specific interval we are looking for probabilities for
l <- 15 # How many we expect in an hour
ppois(q = x, l)
```

    So, based on this, the probability of hearing 9 or more titi calls is 56.8%

2. *What is the probability that she will hear no calls in a session?*

    Not sure how this one will work? But let's try using the same formula:

```{r}
x <- 0
l <- 15
ppois(q = 0, l)
```

*Yikes!* So basically a **very** small chance of hearing 0 . That makes sense!
    
3. *What is the probability that she will hear exactly 3 calls in a session?*

    Well, it seemed to work with setting it for just ) in the last question, let's see what happens when we set x
    at 3 and not a range like in question 1.
    
```{r}
x <- 3
l <- 15
ppois(q = x, l)
```

Ok, still a *really* number. While if the average is 15 and 3 is way below that, so a small probability for
exactly 3 makes sense, I am not sure if this is how we sent up the function for this type of question.

4. *Plot the relevant Poisson mass function over the values in range 0 ≤ x ≤ 30.*

    Ok, here is where it starts getting tricky! Let's see what we can do:

```{r}
x <- 0:30
l <- 15
barplot(dpois(x=x,l),names.arg=x,space=0,xlab="Number of Calls",ylab="Probability of Call
Frequency",main="Probability of Titi Monkey Call Frequencies")
```

WHAT?! Look at that! It kind of makes sense too, that the distribution for highest call frequency probability
would be around 15, which is the average, and then taper off towards either end. I hope this is correct... or at
least some version of correct!  


    