---
title: "Module 8"
author: "Laura Brubaker-Wittman"
date: "9/24/2019"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Probabilities and Distributions

## Objectives

The objective of this module is to begin our discussion of statistical inference from a frequentist/classical
statistics approach. Doing so means that we need to cover basics of probability and distributions.

## Recap of Important Terms and Concepts

* **Population** - includes **all** of the elements from a set of data = ***N***
* **Sample** - one or more observations from a population = ***n***
* **Parameter** - a measurable characteristic of a *population*
* **Statistic** - a measurable characteristic of a *sample*

When we do **statistical inference** we are basically trying to draw conclusions about a *population* based on
measurements from a noisy *sample* or trying to evaluate whether it is reasonable to assume that our sample is drawn from a particular population.

This process of trying to draw conclusions is complicated by the fact that…

* Our sample may be *biased*, *non-random*, or *non-representantive* in some way
* There may be unknown or unobserved variables that impact how the sample is related to the population
* The assumptions we make about the population that our sample is drawn from might not be correct

***All of these are common in primatological data and should be carefully considered when performing analyses and
choosing statistical tests to run!***

## Probability

The term **probability** is applied to **population level** variables that describe the magnitude of chance
associated with particular observations or event. Probabilities summarize the relative frequencies of possible
outcomes.Probabilities are properties of **distributions**. Probabilities vary between zero and one. Outcomes that
are impossible have *Pr = 0*, those that are certain have *Pr = 1*.

Example: if we roll a (fair) die, there are 6 possible outcomes, each has a probability of occurring of 1 in 6.
This is referred to as a *frequentist* or *classical* way of thinking about the probability of different outcomes:
the relative frequency with which an event occurs over numerous identical, objective trials.

**Example: Rolling a Die**

We will use the {manipulate} package and the sample() function to explore the effects of sample size on estimates
of the probability of different outcomes. The probability of each outcome (rolling a “1”, “2”,…, “6”) is 1 in 6,
but our estimate of the probability of each possible outcome will change with sample size.

*Here, I ran the code chunk provided in the module, including installing and loading the {manipulate} package.
The code below not only runs the probability of rolling each side of the die randomly (so will be 
different each time you do it) over 100 trials, but also created a histogram of the results. The {manipulate}
package will also let you manipulate how many roles within R Studio without writing new code. A box pops up with a
slider function when you run the code in the console and you can adjust the "n" in real time. This is the part of
the code that reads (so you are saying you want to make a slider for an "n" from 100 to 1000, by increments of
100)*:

**n=slider(0, 10000, initial=100, step=100)**

```{r eval=FALSE, include=FALSE}
library(manipulate)
outcomes <- c(1,2,3,4,5,6)
manipulate(
    hist(sample(outcomes, n, replace = TRUE),
             breaks=c(0.5,1.5,2.5,3.5,4.5,5.5,6.5),
             probability = TRUE,
             main = paste("Histogram of Outcomes of ", n, " Die Rolls", sep=""),
             xlab = "roll",
             ylab = "probability"),
    n=slider(0, 10000, initial=100, step=100)
)
```

    library(manipulate)
    outcomes <- c(1,2,3,4,5,6)
    manipulate(
        hist(sample(outcomes, n, replace = TRUE),
             breaks=c(0.5,1.5,2.5,3.5,4.5,5.5,6.5),
             probability = TRUE,
             main = paste("Histogram of Outcomes of ", n, " Die Rolls", sep=""),
             xlab = "roll",
             ylab = "probability"),
        n=slider(0, 10000, initial=100, step=100)
    )
    
*Note: I was unable to run this code in R Markdown html rendering because manipulate() must be run in R Studio. I am assuming because of the slider code.*

## *CHALLENGES*

Write a function to simulate rolling a die where you pass the number of rolls as an argument. Then, use your
function to simulate rolling two dice 1000 times and take the sum of the rolls. Plot a histogram of those
results.


```{r}
nrolls <- 1000
roll <- function(x) {sample(1:6, x, replace = TRUE)}
two_dice <- roll(nrolls)+roll(nrolls)
hist(two_dice,breaks=c(1.5:12.5), probability = TRUE, main="Rolling Two Dice", xlab = "sum of rolls", ylab = "probability")
```

# Rules of Probability

1. *Pr*(+) = Probability that something occurs = 1

2. *Pr*(∅) = Probability that nothing occurs = 0

3. *Pr*(***A***) = Probability that a particular event ***A*** occurs

    0 ≤ *Pr*(***A***) ≤ 1 *somewhere between 0 and 1, meaning 0, as described means the event does not happen,
    while 1 means it happens all the time, and a decimal between 0 and 1 would represent the precentage of how
    often it happens*
    
4. 